### 2015-03-28 Kokkos_PG_MachineModel.tex
Adds figures for execution spaces and memory spaces in Kokkos.

@@

@@

@@

Memory consistency models are a complex topic in and of themselves
and usually rely on complex operations associated with hardware
caches (for more information see reference~\cite{handp_hardware}. 
Kokkos itself does not {\em require} caches to be present
in hardware and so assumes an extremely weak memory consistency
model. In this model, the programmer should not assume any
specific ordering of memory operations being isused by a
kernel. This has the potential to create race conditions between
memory operations if these are not appropriately protected. In
order to provide a guaranteed that memory are completed, Kokkos
provides a {\em fence} operation which forces the compute 
engine to complete all outstanding memory operations before
any new ones can be issued. With appropriate use of fences,
programmers are thereby able to ensure that guarantees can
be made as to when data will {\em definitely} have been 
written to memory.

@@

\section{Memory consistency}\label{S:Model:Consistency}

A \emph{memory consistency model} describes the behavior of a memory space $M$,
with respect to multiple threads running in an execution space $E$
that access the same data in $M$ concurrently.
One may also encounter this idea as the phrase ``cache coherency.''
We prefer ``memory consistency,'' because Kokkos does not require caches.
Consistency models are a complicated subject which we cannot cover in sufficient detail here.
We encourage readers to refer to standard computer hardware textbooks,
like ``Computer Architecture: A Quantitative Approach'' (Hennessy and Patterson, 2011).

Kokkos imposes a very weak consistency model across an entire execution space.
For example, it does \emph{not} assume sequential consistency.
Much like MPI 3 one-sided communication
or other partitioned global address space (PGAS) programming models,
Kokkos requires an execution space to issue a \emph{memory fence}
to a given memory space, 
in order to bring that memory space into a consistent state.
The fence forces completion of all outstanding memory writes,
and reorders all loads to occur after all those writes complete.

\subsection{Atomic updates}\label{SS:Model:Consistency:Atomic}

Some execution spaces may implement \emph{atomic updates} with respect to some memory spaces.
An atomic update runs on an execution space, and changes a word of data in a memory space.
The atomic update behaves atomically with respect to other atomic updates happening concurrently by that execution space on that memory space.
``Behaves atomically'' means that all of the atomic updates will complete,
and the word of data will be changed in a way that respects some order of those update operations.
Kokkos does not promise any particular order.
Furthermore, for a nonatomic memory read or write to ``see'' the results of atomic updates,
the execution space must first execute a \lstinline!fence()! operation to that memory space.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/kokkos-execution-space.pdf}
\caption{Example Execution Spaces in a Future Computing Node}
\label{fig:kokkos_execution_spaces}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/kokkos-memory-space.pdf}
\caption{Example Memory Spaces in a Future Computing Node}
\label{fig:kokkos_memory_spaces}
\end{center}
\end{figure}

@@

In cases where multiple executing threads attempt to read
a memory address, complete a computation on the item and
write it back to same address in memory, an ordering
collision may occur. These situations, known as
{\em race conditions} (because the data value
stored in memory are the threads complete is dependent
on which thread completes its memory operation last),
are often the cause of non-determinism in parallel
programs. A number of methods can be employed to ensure
that race conditions do not occur in parallel programs
including the use of locks (which allow only a single
thread to gain access to data structure at a time),
critical regions (which allow only one thread to execute
a code sequence at any point in time) and {\em atomic}
operations. Memory operations which are atomic guarantee
that a read, simple computation, and write to memory
are completed a single unit. This might allow application
programmers to safely increment a memory value for instance,
or, more commonly, to safely accumulate values from multiple
threads into a single memory locations.

Although atomic operations are logically an operation
on memory, they are performed by compute units and so
are a property of execution spaces in the Kokkos framework.
Execution spaces are not guaranteed to provide atomic
operations support but attempts to use them on spaces
which do not support atomic access will result in 
a compile time error.

@@

Memory consistency models are a complex topic in and of themselves
and usually rely on complex operations associated with hardware
caches or memory access coherency (for more information 
see reference~\cite{handp_hardware}. 
Kokkos, does not {\em require} caches to be present
in hardware and so assumes an extremely weak memory consistency
model. In the Kokkos model, the programmer should not assume any
specific ordering of memory operations being isused by a
kernel. This has the potential to create race conditions between
memory operations if these are not appropriately protected. In
order to provide a guaranteed that memory are completed, Kokkos
provides a {\em fence} operation which forces the compute 
engine to complete all outstanding memory operations before
any new ones can be issued. With appropriate use of fences,
programmers are thereby able to ensure that guarantees can
be made as to when data will {\em definitely} have been 
written to memory.

@@
